---
title: "Introduction to NorseResearch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to NorseResearch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette shows a practical, end-to-end workflow for using the `NorseResearch` R package with Norse Feedback (NF) version 3 and higher. It also summarizes the major functions so you can find the right tool quickly.

```{r}
library(NorseResearch)
library(tidyverse)
```

## Major Functions At A Glance

Data import and encoding repair:

- `read.csv_nf3()`, `read.csv2_nf3()`, `read_excel_nf3()` for NF3+ exports
- `read_password_protected_excel()` for protected files
- `fix_failed_encoding()` for likely mojibake in character columns

Cleaning and reshaping:

- `clean_NF_names()` standardizes NF export variable names
- `collapse_measures_wide()` collapses `_A#` suffix columns and widens multi-measure items
- `combine_suffix_variables()` and `combine_q_vars()` merge related columns
- `replace_98s_99s()` replaces sentinel values with `NA`
- `nicer_id_var()` creates a more readable ID column
- `get_first_obs()` selects the first observation per respondent

Versioning, lookup, and metadata:

- `check_version_nf()` detects NF versions in a data set
- `lookup_item()`, `get_item_text()` inspect items
- `get_nicer_name()` and `get_nf3_nicer_name()` convert short names to readable names
- `lookup_trigger()`, `find_trigger()`, `find_trigger_among()` locate trigger items

Scoring:

- `score_all()` scores all NF3+ scales in raw scale form
- `score_all_nf3()` specifically targets NF3 scales
- `nf_score()` scores mixed-version NF data
- `score_NORSE_trigger()` keeps subthreshold trigger data
- `score_NORSE_overunder()` and `score_all_NORSE2_ou()` for over-under scoring
- `score_normed_NF()` computes normed scores
- `rename_score_vars()` renames exported `SCORE_*` variables

Analysis and plotting:

- `scale_analysis()`, `scale_analysis2()` analyze scale properties
- `scale_plot()` plots annotated scale histograms
- `item_plot()` plots item distributions
- `trigger_plot()` visualizes trigger item distributions
- `ggplot_icc_plot()` and `ggplot_information_plot()` for IRT plots

Synthetic data generation:

- `random_norse_data()`, `random_assessment_generator()` for mock NF data
- `random_person_generator()`, `random_tx_generator()` for mock person and treatment data

## NF3+ Workflow Example

This example uses a typical NF3+ export file, cleans and collapses the data, scores the NF scales, and then produces a quick scale plot. Replace the file path with your own NF3+ export.

```{r}
# 1) Read NF3+ export data
# raw_nf <- read.csv_nf3("path/to/response_data.csv")
raw_nf <- HF_research_data_2021

# 2) Confirm NF version (should show 3.x or higher)
check_version_nf(raw_nf)

# 3) Fix common encoding issues and standardize names
raw_nf <- fix_failed_encoding(raw_nf)
raw_nf <- clean_NF_names(raw_nf)

# 4) Replace sentinel missing values (98, 99) with NA
raw_nf <- replace_98s_99s(raw_nf)

# 5) Collapse repeated measures and widen multi-measure items
# nf_collapsed <- collapse_measures_wide(raw_nf)

# 6) Score all NF3+ scales
# nf_scored <- score_all(nf_collapsed)

# 7) Optionally compute normed scores
# nf_normed <- score_normed_NF(nf_scored)

# 8) Quick visualization of a scale distribution
scale_plot(raw_nf, item.names = sad.names,title_obj = "sad")
```

## Multi-Export Example (NF3+ Calibration Style)

If you have multiple exports, the common workflow is: read each export, clean and collapse, then combine and score.

```{r, include=FALSE,eval=FALSE}
export_paths <- list(
  "export_01/response_data.csv",
  "export_02/response_data.csv",
  "export_03/response_data.csv"
)

raw_exports <- purrr::map(export_paths, read.csv_nf3)
raw_exports <- purrr::map(raw_exports, fix_failed_encoding)
raw_exports <- purrr::map(raw_exports, clean_NF_names)
raw_exports <- purrr::map(raw_exports, replace_98s_99s)

collapsed <- purrr::map(raw_exports, collapse_measures_wide)

combined <- dplyr::bind_rows(collapsed, .id = "export_id")

combined_scored <- score_all(combined)
```

## Trigger Item Utilities (NF3+)

Trigger items are often used to control whether a scale is considered valid. Use these helpers to inspect and manage trigger items.

```{r}
# Find trigger item for a scale name
lookup_trigger("Depresjon")

# Find trigger among a list of items
find_trigger_among(c("Q25", "Q26", "Q27"))
```

## Scale Analysis Example

```{r}
sa <- scale_analysis2("sad affect", sad.names.nf3, raw_nf)
print(sa)
plot(sa)
```

## Notes For NF3+

- Use `check_version_nf()` early to confirm the export version. If you find mixed versions, `nf_score()` can help.
- Prefer `score_all()` for NF3+ data to ensure all relevant scales are scored consistently.
- If your exports already contain `SCORE_*` variables, use `rename_score_vars()` to map them to consistent names before analysis.

