---
title: "Introduction to NorseResearch"
output:
  rmarkdown::html_vignette:
    math_method: null
    pandoc_args: ["--math-method=mathjax"]
vignette: >
   %\VignetteIndexEntry{Introduction to NorseResearch}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a practical, end-to-end workflow for using the `NorseResearch`
R package with data exported from **Norse Feedback (NF)**, the Norwegian
web-based outcome-measure system used in mental-health and substance-use
treatment. The package reads NF exports, cleans and reshapes them, scores the
standard scales, and helps you inspect and plot the items and scales.

The package contains the code and item documentation, but **not** the patient
data itself, which is confidential. A small set of **synthetic** data
(`synthetic_data`) is bundled so the examples run without a real export.

```{r}
library(NorseResearch)
```

## Functions at a glance

The main functions, grouped by task. Use `?name` in R for full documentation of
any function (for example, `?score_all`).

**Read and repair imports**

- [`read.csv_nf3()`](?read.csv_nf3), [`read.csv2_nf3()`](?read.csv2_nf3), [`read_excel_nf3()`](?read_excel_nf3) for NF3+ exports.
- [`read_password_protected_excel()`](?read_password_protected_excel) for password-protected files.
- [`fix_failed_encoding()`](?fix_failed_encoding) for likely mojibake in character columns.

**Clean and reshape**

- [`clean_NF_names()`](?clean_NF_names) standardizes NF export variable names.
- [`collapse_versioned_columns()`](?collapse_versioned_columns) deterministically collapses numeric suffix versions.
- [`collapse_measures_wide()`](?collapse_measures_wide) collapses `_A#` suffix columns and widens multi-measure items.
- [`combine_suffix_variables()`](?combine_suffix_variables) and [`combine_q_vars()`](?combine_q_vars) merge related columns.
- [`has_no_98s_99s()`](?has_no_98s_99s) checks numeric columns for `-98`/`-99` sentinels before a pipeline continues.
- [`replace_98s_99s()`](?replace_98s_99s) is a generic sentinel-replacement utility; NF scoring uses [`prepare_nf_items()`](?prepare_nf_items) instead.
- [`nicer_id_var()`](?nicer_id_var) creates a readable ID column.
- [`get_first_obs()`](?get_first_obs) and [`get_first_nonmissing_obs()`](?get_first_nonmissing_obs) select respondent observations.

**Versioning, lookup, and metadata**

- [`check_version_nf()`](?check_version_nf) detects which NF version(s) are present.
- [`lookup_item()`](?lookup_item) inspects an item's text by `Q`-number.
- [`get_nicer_name()`](?get_nicer_name) and [`get_nf3_nicer_name()`](?get_nf3_nicer_name) convert short names to readable names.
- [`lookup_trigger()`](?lookup_trigger), [`lookup_trigger_among()`](?lookup_trigger_among) locate trigger items. (The older `find_trigger()` and `find_trigger_among()` are deprecated in favor of these.)

**Score scales**

- [`audit_nf_reverse()`](?audit_nf_reverse) checks whether positive items have already been reversed in an export.
- [`prepare_nf_items()`](?prepare_nf_items) prepares an item-level copy with explicit input coding and sentinel handling.
- [`score_all()`](?score_all) scores all NF2/NF3 scales in raw scale form.
- [`score_all_nf3()`](?score_all_nf3) specifically targets NF3 scales.
- [`nf_score()`](?nf_score) is a deprecated mixed-version compatibility helper; use [`score_all()`](?score_all) for new work.
- [`score_NORSE_trigger()`](?score_NORSE_trigger) keeps subthreshold trigger data.
- [`score_NORSE_overunder()`](?score_NORSE_overunder) and [`score_all_NORSE2_ou()`](?score_all_NORSE2_ou) for over/under scoring.
- [`score_normed_NF()`](?score_normed_NF) computes normed scores.
- [`rename_score_vars()`](?rename_score_vars) renames exported `SCORE_*` variables.

**Analyze and plot**

- [`scale_analysis2()`](?scale_analysis2) analyzes scale properties (reliability and IRT). Use this rather than the deprecated `scale_analysis()`.
- [`scale_plot()`](?scale_plot) plots an annotated scale histogram.
- [`item_plot()`](?item_plot) plots an item distribution.
- [`trigger_plot()`](?trigger_plot) visualizes a trigger-item distribution (needs a `mirt` fit).
- [`ggplot_icc_plot()`](?ggplot_icc_plot) and [`ggplot_information_plot()`](?ggplot_information_plot) for IRT plots.

**Generate mock NF data**

- [`random_norse_data()`](?random_norse_data) for mock NF data; [`random_person_generator()`](?random_person_generator) and [`random_tx_generator()`](?random_tx_generator) build the underlying person and treatment layers.

## A quick start with the bundled data

The bundled `synthetic_data` is a reproducible NF2 data set made with the
current generator. It follows NF2 trigger rules, so items after a scale's
trigger are `NA` when the scale is not administered. It also includes
pre-computed scale columns such as `sad` and `cog`, making it the quickest way
to exercise the lookup, analysis, and plotting functions.

```{r}
data(synthetic_data)
synthetic_data
```

```{r}
# 1) Which NF version(s) are present?
check_version_nf(synthetic_data)
```

```{r}
# 2) Readable names: short code -> presentation name
get_nf3_nicer_name("sad")       # "Sad Affect"

# Item text by Q-number
lookup_item("Q42", version = 2) # "I am feeling depressed"
```

```{r}
# 3) Trigger items
lookup_trigger("Somatic")              # "Q51"
lookup_trigger_among(somAnx.names)     # "Q51"
```

## Analyzing and plotting a scale

`scale_analysis2()` returns an object with descriptive statistics, reliability
coefficients (omega and alpha), and IRT information. It takes a character vector
of item names, not the scale code, so pass one of the bundled name vectors such
as `sad.names`.

```{r}
sa <- scale_analysis2("Sad affect", sad.names, synthetic_data,
                      IRTpackage = "mirt", print.now = FALSE)
print(sa)
plot(sa)
```

Plot a scale-score distribution or a single item's responses directly:

```{r}
scale_plot(synthetic_data, item.names = sad.names, title_obj = "Sad Affect")
item_plot("Q42", data = synthetic_data)
```

## Reshaping an export with multiple M-coded measures

NF exports can contain one row per measure code, with item responses stored in
columns such as `M100_Q10` and `M200_Q20`. The following small example
mimics that structure. It has two M-coded measures for each respondent/date,
an `_A1` duplicate that must be collapsed, and a `Q11_1` paired item that
must be combined after widening.

```{r}
raw_measures <- data.frame(
  Respondent_ID = c("p01", "p01", "p02", "p02"),
  Submitted = as.Date(c("2026-01-02", "2026-01-02",
                        "2026-01-05", "2026-01-05")),
  Short_code = c("M100", "M200", "M100", "M200"),
  M100_Q10 = c(NA, NA, 3, NA),
  M100_Q10_A1 = c("1", NA, NA, NA),
  M100_Q11 = c(NA, NA, 4, NA),
  M100_Q11_1 = c(2, NA, NA, NA),
  M200_Q20 = c(NA, 3, NA, 5),
  M200_Q21 = c(NA, 4, NA, 6),
  check.names = FALSE
)

# The primitive orders numeric suffixes and converts the numeric-like
# character duplicate to numeric. collapse_measures_wide() calls this
# internally for the same _A# step.
collapsed_columns <- collapse_versioned_columns(
  raw_measures, suffix_pattern = "_A\\d+$", coerce = "common"
)
collapsed_columns[c("Short_code", "M100_Q10", "M100_Q11")]
```

Now widen the measure columns. `collapse_measures_wide()` uses `Short_code`
to select the matching `M####_` columns and returns one row per respondent,
date, and measure code. It also delegates the final `Q#_1` merge to the same
suffix primitive.

```{r}
wide_by_measure <- collapse_measures_wide(
  collapsed_columns,
  item_keep_pattern = "^Q\\d+$",
  coerce_wide_numeric = TRUE
)
wide_by_measure
```

If the analysis needs one row per respondent/date rather than one row per
M-coded measure, consolidate the measure rows after widening. This explicit
aggregator preserves missingness and stops if two measures provide different
values for the same Q item.

```{r}
first_nonmissing_checked <- function(x) {
  values <- x[!is.na(x)]
  if (length(unique(values)) > 1L) {
    stop("Conflicting non-missing values for one respondent/date.")
  }
  if (length(values)) values[[1L]] else NA_real_
}

combined_items <- wide_by_measure |>
  dplyr::group_by(Respondent_ID, Submitted) |>
  dplyr::summarise(
    dplyr::across(dplyr::starts_with("Q"), first_nonmissing_checked),
    .groups = "drop"
  )
combined_items
```

For a real export, use `collapse_measures_wide(raw_nf)` directly: it performs
the `_A#` collapse and `Q#_1` merge internally. Call
`collapse_versioned_columns()` separately when you need to inspect or stage
the column collapse, or when applying the same deterministic suffix policy to
another set of exported columns.

## Recommended export-to-score workflow

The following is the standard path for an NF3.1 export. Do not replace the
special codes before the audit: `read.csv_nf3()` may import `-99` as `NA`, while
`-98` must remain available for the audit's sensitivity check. The scoring
function then handles these values according to the item and scale policy.

```{r recommended-scoring-workflow, eval = FALSE}
# 1. Read and clean the export.
raw_nf <- read.csv_nf3("path/to/response_data.csv")
items <- fix_failed_encoding(raw_nf)
items <- clean_NF_names(items)
items <- collapse_measures_wide(items)

# 2. Confirm the version and audit the response direction.
check_version_nf(items)
verdict <- audit_nf_reverse(items)

if (is.na(verdict) || verdict == "mixed") {
  stop("Review the audit with verbose = TRUE before scoring.")
}

# 3. Translate the audit verdict into the explicit scoring convention.
input_coding <- if (verdict == "consistent") {
  "higher_is_worse"
} else {
  "agreement"
}

# 4. Score the scales. Original item columns remain unchanged.
scored <- score_all(items, input_coding = input_coding)
```

For repeated assessments, add the export's identifier and date columns to the
audit, for example `patient_id = "Respondent_ID"` and
`order_by = "Submitted"`. The audit then uses one assessment per patient for
the correlation check while `score_all()` still scores every row.

The main symptom and resource scores in `scored` use higher = more problems.
Alliance, Therapy Preferences/Needs, QOL, and Norse are exceptions: their
ordinary responses are not reversed, both `-98` and `-99` become missing for
scoring, and their original meaning is retained. QOL item Q226 remains on its
0--10 response scale.

## Understanding the audit result

An NF export may contain original agreement ratings or values already converted
to **higher = more problems**. Use `audit_nf_reverse()` on numeric item columns
while the original `-98` and `-99` codes are still present. It compares both
interpretations without changing the data.

This reproducible example uses the bundled NF3.1 item definitions for
Impulsivity (mixed wording), Social Support (all positive wording), and the
negative-only Worry and Sad Affect anchors. The responses are synthetic and
intentionally correlated to illustrate the diagnostic; they are not patient
data or a model of typical NF correlations.

```{r audit-example-data}
audit_scales <- c("Impulsivity", "Social Support", "Worry", "Sad Affect")
audit_metadata <- subset(NF3.1_items, scale_e %in% audit_scales)
positive_items <- audit_metadata$item[trimws(audit_metadata$reverse) == "R"]

set.seed(20260926)
problem_level <- rep(1:7, each = 20)
original_ratings <- as.data.frame(setNames(
  lapply(audit_metadata$item, function(item) {
    response <- pmin(7, pmax(1, problem_level +
      sample(-1:1, length(problem_level), replace = TRUE,
             prob = c(0.15, 0.70, 0.15))))
    if (item %in% positive_items) 8 - response else response
  }),
  audit_metadata$item
))

# Retain the special export codes for the audit to interpret.
original_ratings[1:5, ] <- -98
original_ratings[6, ] <- -99

# Simulate a second export whose positive items were already reversed.
already_reversed <- original_ratings
for (item in positive_items) {
  ordinary <- already_reversed[[item]] %in% 1:7
  already_reversed[[item]][ordinary] <-
    8 - already_reversed[[item]][ordinary]
}

audit_exports <- rbind(
  transform(original_ratings, export_id = "original_agreement"),
  transform(already_reversed, export_id = "already_reversed")
)
```

The default output is a single answer about whether higher numbers consistently
mean more problems:

```{r audit-example-results}
audit_nf_reverse(original_ratings, metadata = audit_metadata)
# "not consistent": positive items still use original agreement ratings

audit_nf_reverse(already_reversed, metadata = audit_metadata)
# "consistent": positive items already have higher = worse coding

# Keep different exports separate when calculating their correlations.
audit_nf_reverse(audit_exports, metadata = audit_metadata,
                 group_vars = "export_id")
# "mixed": the two exports use different conventions
```

`"not consistent"` here means that the tested positive items need reversal;
it does not mean that some exports have already been reversed and others have
not. That combination gives `"mixed"`, as do conflicting item directions within
a scale. If the evidence is too weak or unresolved, the function returns `NA`
with a brief warning instead of guessing. Established mixed coding takes
precedence over unresolved evidence elsewhere.

The audit handles `-99` as missing. It compares results excluding `-98` with a
sensitivity analysis that recodes `-98` to `1` in higher = worse units. A
positive item's no-problem value on the original agreement scale would be `7`,
which also becomes `1` after reversal. Keep these special codes intact until
after the audit.

### Requesting item-level details

Use `verbose = TRUE` only when you want to inspect the evidence. The returned
list includes the same overall verdict in `status`, a short explanation table
in `issues`, and an item table in `items`. When the default result is `NA`, its
warning reports these issues directly; rerun with `verbose = TRUE` to retain
them for inspection.

```{r audit-example-evidence}
audit_details <- audit_nf_reverse(
  audit_exports, metadata = audit_metadata,
  group_vars = "export_id", verbose = TRUE
)
audit_details$status
audit_details$issues

# Inspect one positive item from each of the two scales being checked.
subset(audit_details$items, item %in% c("Q223", "Q204"),
       select = c(group, scale, item, status, action, item_conflict))
```

Group 1 is the original-agreement export and group 2 is the already-reversed
export. `REVERSE` proposes `8 - response` for ordinary 1--7 responses; `KEEP`
proposes leaving them unchanged. `UNRESOLVED` means the evidence cannot support
an action. The function never modifies the data. Item statuses inherit the
scale's evidence, with conflicts flagged separately; negative items are marked
`"assumed consistent"` because their direction is the trusted starting point.

Additional detail is available in `audit_details$correlations`,
`audit_details$classifications`, and `audit_details$counts`. For example, to
inspect all pair and item-rest comparisons involving Q223:

```{r audit-item-correlations, eval = FALSE}
subset(audit_details$correlations, left == "Q223" | right == "Q223")
```

The verdict summarizes the available checks, not verification of every item.
By default these check mixed-wording scales and Social Support against Worry
and Sad Affect. Other entirely positive scales require justified relationships
in `positive_scale_anchors`; otherwise their items are `"not assessed"`.
Absent scales do not vote, and insufficient evidence in an included scale
prevents a uniform verdict. Thresholds are diagnostic defaults, not validated
NF cutoffs. See `?audit_nf_reverse` for the full decision rules.

Scale and item-rest means use one available item by default, which accommodates
sparse and trigger-driven NF rows without imputing responses. Set `min_items`
or `min_fraction` only when the audit should require stricter completeness.

### How scoring uses the audit result

Once the export convention has been reviewed, supply it explicitly to the scorer:
`"consistent"` corresponds to `input_coding = "higher_is_worse"`, while
`"not consistent"` corresponds to `input_coding = "agreement"`. A `"mixed"` or
`NA` result requires review of batch/item rules before scoring. The audit does
not certify the direction of unassessed items.

The two synthetic exports above describe the same responses. Their scores
should therefore agree after preparation:

```{r audit-to-scoring}
scored_agreement <- score_all(
  transform(original_ratings, Ver_10 = "3.1"), input_coding = "agreement"
)
scored_standardized <- score_all(
  transform(already_reversed, Ver_10 = "3.1"), input_coding = "higher_is_worse"
)
problem_scores <- c("impulsivity", "socSup", "worry", "sad")
isTRUE(all.equal(scored_agreement[problem_scores],
                 scored_standardized[problem_scores]))
```

Problem and resource-deficit scales now have higher = more problems. Positive
resource labels such as Social Support describe **lack of support** when scored
in this direction. **Alliance, Therapy Preferences, QOL, and Norse (Q148,
including Q148.2) are exceptions:** ordinary responses are never reversed, both
`-98` and `-99` are missing, and these outputs retain their original meaning.
QOL remains on its 0--10 scale. The audit excludes these exceptions from its
problem-direction verdict.

The scorer prepares an internal copy and preserves the original item columns
in its output. If you need prepared items for item-level analysis or a
low-level scorer, request them explicitly:

```{r prepare-items-example}
prepare_nf_items(
  data.frame(Q223 = c(7, -98, -99), Q226 = c(7, -98, -99),
             Q148 = c(7, -98, -99)),
  input_coding = "agreement", version = "3"
)
```

For Q223, the results are `1, 1, NA`; for Q226 and Q148, they are `7, NA, NA`.
Never reverse the special codes themselves or reverse an item twice. When
passing already prepared items to a scorer, use `input_coding = "higher_is_worse"`.
The scoring wrappers use that convention by default for backward compatibility;
it is an assumption, not an automatic audit.

## Generating your own mock data

If you would rather not load the bundled data, generate a fresh mock NF data
set. By default it is NF2 with trigger-driven `NA` values for unopened scales.

```{r}
mock <- random_norse_data(20, num_obs = 5)
check_version_nf(mock)
item_plot("Q42", data = mock)
```

### Mixed NF2/NF3 mock exports

Use `versions = c("2", "3")` to create a mixed export. The generated `Ver_10`
column identifies the version for each row. Within every patient, NF2 rows
always come before NF3 rows; a patient may have only one version if their course
falls entirely before or after the rollout. Set `include_98` and/or
`include_99` to simulate the raw sentinel values. `sentinel_probability` is the
probability that an administered item is replaced by one of the requested
sentinels.

```{r}
set.seed(20260903)
mixed_mock <- random_norse_data(
  20,
  num_obs = 5,
  versions = c("2", "3"),
  include_98 = TRUE,
  include_99 = TRUE,
  sentinel_probability = 0.05
)

check_version_nf(mixed_mock)
dplyr::count(mixed_mock, Ver_10)
```

For NF3, `nf3_trigger_threshold` controls the simulated opening threshold
(default `4.5`); NF2 instead uses the documented thresholds in `nf2.1.logic`.
`score_all()` handles these sentinel values automatically (including its
exceptions for Alliance, Preferences, QOL, and Norse). Use `prepare_nf_items()`
for item-level scoring preparation. `replace_98s_99s()` is a generic utility
that does not implement these exceptions or handle item direction. The mock
responses are random; they demonstrate the API, not evidence of coherent
clinical correlations or an export's coding convention.

## Notes for NF3+

- Use `check_version_nf()` early to confirm the export version. `score_all()`
  supports mixed NF2/NF3 data; `nf_score()` is retained only for compatibility.
- Prefer `score_all()` for NF3+ data to ensure all relevant scales are scored
  consistently.
- If your exports already contain `SCORE_*` variables, use
  `rename_score_vars()` to map them to consistent names before analysis.

## Next steps

- `?NorseResearch` — package overview and the list of bundled data objects.
- The [reference index](https://github.com/andrewmcaleavey/NorseResearch)
  lists every exported function and dataset.
