---
title: "Introduction to NorseResearch"
output:
  rmarkdown::html_vignette:
    math_method: null
    pandoc_args: ["--math-method=mathjax"]
vignette: >
   %\VignetteIndexEntry{Introduction to NorseResearch}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a practical, end-to-end workflow for using the `NorseResearch`
R package with data exported from **Norse Feedback (NF)**, the Norwegian
web-based outcome-measure system used in mental-health and substance-use
treatment. The package reads NF exports, cleans and reshapes them, scores the
standard scales, and helps you inspect and plot the items and scales.

The package contains the code and item documentation, but **not** the patient
data itself, which is confidential. A small set of **synthetic** data
(`synthetic_data`) is bundled so the examples run without a real export.

```{r}
library(NorseResearch)
```

## Functions at a glance

The main functions, grouped by task. Use `?name` in R for full documentation of
any function (for example, `?score_all`).

**Read and repair imports**

- [`read.csv_nf3()`](?read.csv_nf3), [`read.csv2_nf3()`](?read.csv2_nf3), [`read_excel_nf3()`](?read_excel_nf3) for NF3+ exports.
- [`read_password_protected_excel()`](?read_password_protected_excel) for password-protected files.
- [`fix_failed_encoding()`](?fix_failed_encoding) for likely mojibake in character columns.

**Clean and reshape**

- [`clean_NF_names()`](?clean_NF_names) standardizes NF export variable names.
- [`collapse_measures_wide()`](?collapse_measures_wide) collapses `_A#` suffix columns and widens multi-measure items.
- [`combine_suffix_variables()`](?combine_suffix_variables) and [`combine_q_vars()`](?combine_q_vars) merge related columns.
- [`replace_98s_99s()`](?replace_98s_99s) replaces sentinel `-98`/`-99` with `1`/`NA`.
- [`nicer_id_var()`](?nicer_id_var) creates a readable ID column.
- [`get_first_obs()`](?get_first_obs) selects the first observation per respondent.

**Versioning, lookup, and metadata**

- [`check_version_nf()`](?check_version_nf) detects which NF version(s) are present.
- [`lookup_item()`](?lookup_item) inspects an item's text by `Q`-number.
- [`get_nicer_name()`](?get_nicer_name) and [`get_nf3_nicer_name()`](?get_nf3_nicer_name) convert short names to readable names.
- [`lookup_trigger()`](?lookup_trigger), [`lookup_trigger_among()`](?lookup_trigger_among) locate trigger items. (The older `find_trigger()` and `find_trigger_among()` are deprecated in favor of these.)

**Score scales**

- [`score_all()`](?score_all) scores all NF2/NF3 scales in raw scale form.
- [`score_all_nf3()`](?score_all_nf3) specifically targets NF3 scales.
- [`nf_score()`](?nf_score) is a deprecated mixed-version compatibility helper; use [`score_all()`](?score_all) for new work.
- [`score_NORSE_trigger()`](?score_NORSE_trigger) keeps subthreshold trigger data.
- [`score_NORSE_overunder()`](?score_NORSE_overunder) and [`score_all_NORSE2_ou()`](?score_all_NORSE2_ou) for over/under scoring.
- [`score_normed_NF()`](?score_normed_NF) computes normed scores.
- [`rename_score_vars()`](?rename_score_vars) renames exported `SCORE_*` variables.

**Analyze and plot**

- [`scale_analysis2()`](?scale_analysis2) analyzes scale properties (reliability and IRT). Use this rather than the deprecated `scale_analysis()`.
- [`scale_plot()`](?scale_plot) plots an annotated scale histogram.
- [`item_plot()`](?item_plot) plots an item distribution.
- [`trigger_plot()`](?trigger_plot) visualizes a trigger-item distribution (needs a `mirt` fit).
- [`ggplot_icc_plot()`](?ggplot_icc_plot) and [`ggplot_information_plot()`](?ggplot_information_plot) for IRT plots.

**Generate mock NF data**

- [`random_norse_data()`](?random_norse_data) for mock NF data; [`random_person_generator()`](?random_person_generator) and [`random_tx_generator()`](?random_tx_generator) build the underlying person and treatment layers.

## A quick start with the bundled data

The bundled `synthetic_data` is a reproducible NF2 data set made with the
current generator. It follows NF2 trigger rules, so items after a scale's
trigger are `NA` when the scale is not administered. It also includes
pre-computed scale columns such as `sad` and `cog`, making it the quickest way
to exercise the lookup, analysis, and plotting functions.

```{r}
data(synthetic_data)
synthetic_data
```

```{r}
# 1) Which NF version(s) are present?
check_version_nf(synthetic_data)
```

```{r}
# 2) Readable names: short code -> presentation name
get_nf3_nicer_name("sad")       # "Sad Affect"

# Item text by Q-number
lookup_item("Q42", version = 2) # "I am feeling depressed"
```

```{r}
# 3) Trigger items
lookup_trigger("Somatic")              # "Q51"
lookup_trigger_among(somAnx.names)     # "Q51"
```

## Analyzing and plotting a scale

`scale_analysis2()` returns an object with descriptive statistics, reliability
coefficients (omega and alpha), and IRT information. It takes a character vector
of item names, not the scale code, so pass one of the bundled name vectors such
as `sad.names`.

```{r}
sa <- scale_analysis2("Sad affect", sad.names, synthetic_data,
                      IRTpackage = "mirt", print.now = FALSE)
print(sa)
plot(sa)
```

Plot a scale-score distribution or a single item's responses directly:

```{r}
scale_plot(synthetic_data, item.names = sad.names, title_obj = "Sad Affect")
item_plot("Q42", data = synthetic_data)
```

## Scoring a real NF export

The scoring functions (`score_all()` and `score_all_nf3()`) expect a
clean data set whose item columns are already reduced to short `Q#` names. They
are built around real NF exports, which are larger and confidential; the bundled
`synthetic_data` is not the right input for them. A typical NF3+ workflow looks
like this — replace the file path with your own export.

```{r, eval = FALSE}
# 1) Read the NF3+ export
raw_nf <- read.csv_nf3("path/to/response_data.csv")

# 2) Confirm the version (3.x or higher)
check_version_nf(raw_nf)

# 3) Fix encoding and standardize names
raw_nf <- fix_failed_encoding(raw_nf)
raw_nf <- clean_NF_names(raw_nf)

# 4) Collapse repeated measures and widen multi-measure items
nf_collapsed <- collapse_measures_wide(raw_nf)

# 5) Score all NF3+ scales. score_all() handles -98/-99 automatically.
nf_scored <- score_all(nf_collapsed)

# 6) Optionally compute normed scores
nf_normed <- score_normed_NF(nf_scored)
```

### Multiple exports

When you have several exports, read and clean each one, then combine and score.

```{r, eval = FALSE}
export_paths <- list(
  "export_01/response_data.csv",
  "export_02/response_data.csv",
  "export_03/response_data.csv"
)

raw_exports <- purrr::map(export_paths, read.csv_nf3)
raw_exports <- purrr::map(raw_exports, fix_failed_encoding)
raw_exports <- purrr::map(raw_exports, clean_NF_names)

collapsed <- purrr::map(raw_exports, collapse_measures_wide)

combined <- dplyr::bind_rows(collapsed, .id = "export_id")

combined_scored <- score_all(combined)
```

## Generating your own mock data

If you would rather not load the bundled data, generate a fresh mock NF data
set. By default it is NF2 with trigger-driven `NA` values for unopened scales.

```{r}
mock <- random_norse_data(20, num_obs = 5)
check_version_nf(mock)
item_plot("Q42", data = mock)
```

### Mixed NF2/NF3 mock exports

Use `versions = c("2", "3")` to create a mixed export. The generated `Ver_10`
column identifies the version for each row. Within every patient, NF2 rows
always come before NF3 rows; a patient may have only one version if their course
falls entirely before or after the rollout. Set `include_98` and/or
`include_99` to simulate the raw sentinel values. `sentinel_probability` is the
probability that an administered item is replaced by one of the requested
sentinels.

```{r}
set.seed(20260903)
mixed_mock <- random_norse_data(
  20,
  num_obs = 5,
  versions = c("2", "3"),
  include_98 = TRUE,
  include_99 = TRUE,
  sentinel_probability = 0.05
)

check_version_nf(mixed_mock)
dplyr::count(mixed_mock, Ver_10)
```

For NF3, `nf3_trigger_threshold` controls the simulated opening threshold
(default `4.5`); NF2 instead uses the documented thresholds in `nf2.1.logic`.
`score_all()` handles these sentinel values automatically (including its
special handling of process variables). Use `replace_98s_99s()` only when you
need its general-purpose `-98` to `1` and `-99` to `NA` recoding before another
analysis.

## Notes for NF3+

- Use `check_version_nf()` early to confirm the export version. `score_all()`
  supports mixed NF2/NF3 data; `nf_score()` is retained only for compatibility.
- Prefer `score_all()` for NF3+ data to ensure all relevant scales are scored
  consistently.
- If your exports already contain `SCORE_*` variables, use
  `rename_score_vars()` to map them to consistent names before analysis.

## Next steps

- `?NorseResearch` — package overview and the list of bundled data objects.
- The [reference index](https://github.com/andrewmcaleavey/NorseResearch)
  lists every exported function and dataset.
